Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-26 paper.tex
Yans updates

The Long ARG contains 749 samples designated by Nextclade [CITE] as Pango
recombinants, classed into 33 Pango lineages; this
reduces to 711 samples from the same 33 lineages
when singleton recombinants (i.e. recombination
nodes subtending only one sample) are removed.
In contrast, using the alternative GISAID designation process [CITE], only
515 samples in the Long ARG are labelled as Pango recombinants,
which fall into 38 lineages (when singleton recombinants are excluded this
drops to 514 samples from 35 lineages). The variation in Nextclade versus
GISAID classifications highlights the frequent uncertainty that exists when
assigning Pango X lineages to samples.

@@

\section{Discussion}
% PARA 1. SARS-CoV-2 is still important, and the vast hordes of data provide an
% unprecedented opportunity to study the evolution of a virus. Even if the
% public health emergency is officially over, it doesn't mean it's not coming
% back and we should be prepared. Recombination is now a vital factor, and
% present a way in which it can be incorporated. Our results are accurate.
Although the COVID-19 pandemic is no longer considered a global emergency
by the WHO, the prevalence of SARS-CoV-2 continues to be high worldwide
with persistent high levels of infection. This high prevalence has allowed the
proliferation of many variants, with more than 600 Pango-designated lineages
circulating globally in the last three months (January to March, 2023;
https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
higher risk of coinfection, increasing opportunities for new phenotypically
distinct recombinants to emerge and spread. Phylogenetic approaches have
been central to pandemic responses thus far [TODO: Examples and citations].
However, as recombinant lineages are now endemic and it is clear that
recombination is a major feature of ongoing SARS-CoV-2 evolution,
it is imperative that these methods are updated to incorporate the
effects of recombination, and future public health interventions are
not based on incomplete and potentially biased evolutionary models.
Here we have introduced the first method to infer an evolutionary
history that jointly models genetic inheritance, mutation and
recombination at pandemic scale, and illustrated how this single
structure accurately captures results derived by many different means.

>>>>

The Long ARG contains 749 samples designated by Nextclade [CITATION] as Pango
recombinants, classed into 33 Pango lineages; this
reduces to 711 samples from the same 33 lineages
when singleton recombinants (i.e. recombination
nodes subtending only one sample) are removed.
In contrast, using the alternative GISAID designation process [CITATION], only
515 samples in the Long ARG are labelled as Pango recombinants,
which fall into 38 lineages (when singleton recombinants are excluded this
drops to 514 samples from 35 lineages). The variation in Nextclade versus
GISAID classifications highlights the frequent uncertainty that exists when
assigning Pango X lineages to samples.

@@

\section{Discussion}
% PARA 1. SARS-CoV-2 is still important, and the vast hordes of data provide an
% unprecedented opportunity to study the evolution of a virus. Even if the
% public health emergency is officially over, it doesn't mean it's not coming
% back and we should be prepared. Recombination is now a vital factor, and
% present a way in which it can be incorporated. Our results are accurate.
Although the COVID-19 pandemic is no longer considered a global emergency
by the WHO, the prevalence of SARS-CoV-2 continues to be high worldwide
with persistent high levels of infection. This high prevalence has allowed the
proliferation of many variants, with more than 600 Pango-designated lineages
circulating globally in the last three months (January to March, 2023;
https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
higher risk of coinfection, increasing opportunities for new phenotypically
distinct recombinants to emerge and spread. Phylogenetic approaches have
been central to pandemic responses thus far [FIXME: Examples and citations].
However, as recombinant lineages are now endemic and it is clear that
recombination is a major feature of ongoing SARS-CoV-2 evolution,
it is imperative that these methods are updated to incorporate the
effects of recombination, and future public health interventions are
not based on incomplete and potentially biased evolutionary models.
Here we have introduced the first method to infer an evolutionary
history that jointly models genetic inheritance, mutation and
recombination at pandemic scale, and illustrated how this single
structure accurately captures results derived by many different means.
